A Study of the Effectiveness of Usage Examples in REST API Documentation

### 2017-03-20 controlled_study.tex
down to 8 pages as per guideline

We designed and performed a controlled study with experienced software engineers. Participants were divided into two groups and given the same set of API tasks to complete. One group was given the official WordPress REST API documentation and another group was given a enhanced version of the documentation where three usage examples were added. Using a novel technique, we collected 539 API calls made by the participants. We analyzed the data and observed recurring obstacles faced by the participants while performing the API tasks using the official documentation that lacks usage examples. Our contributions on this paper is as follows:

@@

Starting January 2017, the WordPress REST API V2 is no longer required to be installed as a separate plug-in since it's bundled with every WordPress installation. According to the code repository on GitHub, there are a total of 99 and 46 contributors that had at least one commit to the code repository of the API and its documentation, respectively. These properties satisfy R1, our requirements for using a representative API.

By selecting an open-source project we are able to access the source-code to inspect the implementation and documentation technique of the WordPress REST API. The API implements a self-documenting feature where API developers expose API endpoints over HTTP OPTIONS verb to explain the API elements. To implement this feature, the API developers describe the API elements in the code. For example:

@@

In addition to the auto-generated documentation of the the schema and API actions, custom content is added by the contributors to the API documentation project to provide prosaic overviews and usage examples.

@@

We conducted two pilot studies to evaluate and understand an effective process for performing this study. The first pilot study involved four participants that were invited to join the first author in-person or using a video conferencing software. The study involved tasks using two APIs, the WordPress REST API V2 and the GMail REST API. Each participant was given one of the two APIs and a set of tasks to complete using the API within an hour. We gave the participants an online answer form to record the answers to the tasks. The primary findings from this pilot are as follows: 1) asking participants to use an API to perform the tasks required significant overhead time for them to setup a development environment with the proper API credentials, 2) the intermediate trial attempts of using the API are potentially more valuable than the final answer as it allows us to understand API client developer information needs that may are not answered by the documentation, 3) the number of tasks for the study had to be reduced so the participants could complete the tasks within the one hour limit, and 4) for GMail API, participants used up a large portion of their time on setting up their API credentials that requires understanding of OAuth.

@@

To overcome the shortcomings found from the first pilot study, we decided to develop a web-based REST API explorer as shown on Figure \ref{fig:api_explorer} that allows participants to use their browser to make the API calls without setting up any development environment. The web-based API explorer only requires the inputs I1-4, and displays the HTTP response headers and body on the click of a button. Thus the participants could focus on using the right input and verifying the output without having to write any code. The web-base REST API explorer also allowed us to automatically collect all the trial API calls that the participants make for each API task. A second pilot study involving seven new participants was performed to understand the features of the web-based REST API explorer in practice and to improve the user interface based on feedback from the participants. Participants completed the study on their own without having to meet in-person or over video conferencing. Only the WordPress REST API was used to focus on REST APIs without the required learning curve associated with OAuth. We found encouraging results from this pilot study as the collected data showed patterns of mistakes that API client developers make that can be reduced by adding usage examples in the API documentation.

@@

\subsubsection{Data Analysis}
We analyzed the data artifact qualitatively to understand the obstacles that REST API client developers face when using an API documentation that lacks usage examples. The data artifact exported from the web-based REST API explorer included a table with raw data. For each row, the table contains the following columns: API task, participant identifier, timestamp, API trial number, I1-4, response headers and body. It allowed us to observe and categorize the different input values that the participants used to perform the given API tasks. The values for I1-4 used by the participants were manually coded to group the participant responses into categories. The categories help us determine the types of information that API client developers need in the API documentation to perform API tasks successfully. The analysis started with an empty set of codes and new codes were introduced to describe scenarios that did not fall under  codes that were already applied. The first author applied the codes on the raw data artifact and the co-authors verified the codes randomly. To resolve disagreements, the authors discussed and updated the codes accordingly. The codes were then further categorized to higher level grouping to represent common problems faced by the study participants for each group.

@@

In Figure \ref{graph:productivity} and  \ref{graph:time} we see that, for the simpler tasks T2 and T3, G1 had used more trial API calls and more time on average and reached successful answers with a lower percentage compared to G2. For T4, where the participants are required to inspect a HTTP response header, G2 needed more API calls on average, but also yielded 24\% higher success rate. For the more complicated tasks T5 and T6, we see that participants in G1 had very low success rate and very high number of average API trials compared to G2. The data here clearly depicts that without usage examples REST API client developers spend more time and require more trial and error attempts yet resulting in lower success rates, and the situation gets worse for relatively complicated tasks.

@@

\textbf{Observation 1.} REST API client developers want usage examples along with the definition of API elements. From analyzing the participant feedback, we've observed usage examples mentioned as the most frequent and central topic of interest. 11 of the 16 G2 participants mentioned the lack of examples with HTTP headers, request and response as the primary problem with the API documentation. On the other hand, 7 out of the 10 G2 participants mentioned the examples being helpful from the enhanced documentation. This is also supported by our quantitative analysis as discussed later in this section.

@@

\subsubsection{Measuring API Usability}
Rama et al. presented a set of formulas for computing a measure of API usability based on the API's structural components such as classes, methods, parameters, return values, thread-safety, etc~\cite{rama}. Scheller et al. presented a framework for automatically measuring the complexity of an API~\cite{scheller2015automated}. They identified a list of measurable API properties and provided formulas to compute complexities related to the interfaces, implementation and setup of an API. Grill et al. suggested using a combination of expert opinion and developer workshops to identify and collect feedback about API and its documentation related problems \cite{Grill2012}. Farooq et al. recommended using peer reviews of API code in addition to API usability studies to uncover API related bugs and incorporate feedback~\cite{Farooq:2010:AUP:1753326.1753677}. In our work, instead of measuring overall API usability, we have focused specifically on the relationship of REST API usability with respect to usage examples.

\subsubsection{Controlled Studies}
Nasehi et al. performed a controlled study to understand if API unit tests can be used to provide as usage examples to facilitate API client developers \cite{5609553}. The researchers found that the examples from the unit tests helped understanding the API concepts better but it was challenging for the participants to locate relevant examples from the corpus of unit tests. Endrikat performed a controlled study with four groups of participants to understand the impact of API documentation on APIs that are implemented using programming languages with static and dynamic type systems \cite{Endrikat:2014:ADS:2568225.2568299}. Participants were given a set of failing unit tests and were asked to make them pass by writing code using the studied API. They found that the participant group using a statically typed API with explicit documentation could get the tests to pass in less time than the other groups. Ellis et al. performed a controlled study with two groups of developers to understand the impact of using a constructor vs. a factory method design pattern on API client development time \cite{Ellis:2007:FPA:1248820.1248863}. They found that API client developers needed more time to complete API tasks using factory method compared to using a constructor. Our work in this paper is based on a controlled study and shares part of the setup that were used by the aforementioned API related controlled studies with the following differences: our goal is to understand the obstacles faced by API client developers without usage examples;  we focus on REST APIs instead of a local API; and, the participants are professional software engineers.

>>>>

We designed and performed a controlled study with experienced software engineers. Participants were divided into two groups and given the same set of API tasks to complete. One group was given the official WordPress REST API documentation and another group was given an enhanced version of the documentation where three usage examples were added. Using a novel technique, we collected 539 API calls made by the participants. We analyzed the data and observed recurring obstacles faced by the participants while performing the API tasks using the official documentation that lacks usage examples. Our contributions on this paper is as follows:

@@

Starting January 2017, the WordPress REST API V2 is no longer required to be installed as a separate plug-in since it's bundled with every WordPress installation. According to the code repository on GitHub, there are a total of 99 and 46 contributors that had at least one commit to the code repository of the API and its documentation, respectively. These properties satisfy R1, our requirements for using a representative API. By selecting an open-source project we are able to access the source-code to inspect the implementation and documentation technique of the WordPress REST API. The API implements a self-documenting feature where API developers expose API endpoints over HTTP OPTIONS verb to explain the API elements. To implement this feature, the API developers describe the API elements in the code. For example:

@@

In addition to the auto-generated documentation of the the schema and API actions, contributors add custom content to provide prosaic overviews and usage examples.

@@

We conducted two pilot studies to evaluate and understand an effective process for performing this study. The first pilot study involved four participants that were invited to join the first author in-person or over video conferencing. The study involved tasks using two APIs, the WordPress REST API V2 and the GMail REST API. Each participant was given one of the two APIs and a set of tasks and online answer forms to complete using the API within an hour. The primary findings from this pilot are as follows: 1) participants required significant overhead time to setup a development environment with the proper API credentials, 2) the intermediate trial attempts of using the API are potentially more valuable than the final answer as it allows us to study what obstacles the participants face, 3) the number of tasks for the study had to be reduced to fit within the one hour limit, and 4) for GMail API, participants used up a large portion of their time on learning about how to use  OAuth.

@@

To overcome the shortcomings found from the first pilot study, we decided to develop a web-based REST API explorer as shown on Figure \ref{fig:api_explorer} that allows participants to use their browser to make the API calls without setting up any development environment. The web-based API explorer only requires the inputs I1-4, and displays the HTTP response headers and body on the click of a button. Thus the participants could focus on using the right input and verifying the output without having to write any code. The web-base REST API explorer also allowed us to automatically collect all the trial API calls that the participants make for each API task. A second pilot study involving seven new participants was performed to collect usability related feedback about the web-based REST API explorer. We only used WordPress since it didn't require knowledge about OAuth. From this study, we observed patterns of mistakes that API client developers make that can be reduced by adding usage examples in the API documentation.

@@

\subsubsection{Data Analysis}
We analyzed the data artifact qualitatively to understand the obstacles that REST API client developers face when using an API documentation that lacks usage examples. We exported a table of data from the database behind web-based REST API explorer. For each row, the table contains the following columns: API task, participant identifier, timestamp, API trial number, I1-4, response headers and body. It allowed us to observe and categorize the different input values that the participants used to perform the given API tasks. The values for I1-4 used by the participants were manually coded to group the participant responses into categories. The categories help us determine the types of information that API client developers need in the API documentation to perform API tasks successfully. The analysis started with an empty set of codes and new codes were introduced to describe scenarios that did not fall under  codes that were already applied. The first author applied the codes on the raw data artifact and the co-authors verified the codes randomly. To resolve disagreements, the authors discussed and updated the codes accordingly. The codes were then further categorized to higher level grouping to represent common problems faced by the study participants for each group.

@@

In Figure \ref{graph:productivity} and  \ref{graph:time} we see that, for the simpler tasks T2 and T3, G1 had used more trial API calls and more time on average and reached successful answers with a lower percentage compared to G2. For T4, where the participants are required to inspect an HTTP response header, G2 needed more API calls on average, but also yielded 24\% higher success rate. For the more complicated tasks T5 and T6, we see that participants in G1 had very low success rate and very high number of average API trials compared to G2. The data here clearly depicts that without usage examples REST API client developers spend more time and require more trial and error attempts yet resulting in lower success rates, and the situation gets worse for relatively complicated tasks.

@@

\textbf{Observation 1.} REST API client developers want usage examples along with the definition of API elements. From analyzing the participant feedback, we've observed usage examples mentioned as the most frequent and central topic of interest. 11 of the 16 G1 participants mentioned the lack of examples with HTTP headers, request and response as the primary problem with the API documentation. On the other hand, 7 out of the 10 G2 participants mentioned the examples being helpful from the enhanced documentation. This is also supported by our quantitative analysis as discussed later in this section.

@@

\subsubsection{Measuring API Usability}
Rama et al. presented a set of formulas for computing a measure of API usability based on the API's structural components such as classes, methods, parameters, return values, thread-safety, etc~\cite{rama}. Scheller et al. presented a framework for automatically measuring the complexity of an API~\cite{scheller2015automated}. They identified a list of measurable API properties and provided formulas to compute complexities related to the interfaces, implementation and setup of an API. Grill et al. suggested using a combination of expert opinion and developer workshops to identify and collect feedback about API and its documentation related problems \cite{Grill2012}. Farooq et al. recommended using peer reviews of API code in addition to API usability studies to uncover API related bugs and incorporate feedback~\cite{Farooq:2010:AUP:1753326.1753677}. In our work, instead of measuring API usability, we have focused specifically on the relationship of REST API usability with respect to usage examples.

\subsubsection{Controlled Studies}
Nasehi et al. performed a controlled study to understand if API unit tests can be used to provide as usage examples to facilitate API client developers \cite{5609553}. The researchers found that the examples from the unit tests helped understanding the API concepts better but it was challenging for the participants to locate relevant examples. Endrikat performed a controlled study with four groups of participants to understand the impact of API documentation on APIs that are implemented using programming languages with static and dynamic type systems \cite{Endrikat:2014:ADS:2568225.2568299}. Participants were given a set of failing unit tests and were asked to make them pass by writing code using the studied API. They found that the participant group using a statically typed API with explicit documentation could get the tests to pass in less time than the other groups. Ellis et al. performed a controlled study with two groups of developers to understand the impact of using a constructor vs. a factory method design pattern on API client development time \cite{Ellis:2007:FPA:1248820.1248863}. They found that API client developers needed more time to complete API tasks using factory method compared to using a constructor. Our work in this paper is based on a controlled study and shares part of the setup that were used by the aforementioned controlled studies with the following differences: our goal is to understand the obstacles faced by API client developers without usage examples;  we focus on REST APIs instead of a local API; and, the participants are professional software engineers.
